%%
%% Abstract is MANDATORY. -- Except for MS theses
\begin{abstract}                % Abstract
  Chronic Heart Failure (CHF) is a complex malady caused by numerous disorders of the heart and body resulting in a decreased quality of life. Care for current cardiac patients, post-hospitalization, is not uniform often resulting in relapses and early death. Despite the emergence of numerous therapeutic innovations, prolonging life and shorten hospital stays, CHF patients often have poor prognosis. Monitoring and involving the patient in self-care has been shown to improve upon the prognosis. We propose a new and novel system for cardiac heart failure patient's self-care. This system will collect both physiologic and psychosocial data from the patient using a combination of a custom sensor suite and a smartphone. An exploratory trial evaluates the effectiveness of such an approach.
  
  blank
  
  Review and Comment Received 10/11/2016
  \begin{itemize}
  \item Do not use 1st person e.g. I,me, etc.
  \item get rid of prepositions, starting sentences in particular!
  \item this helps (see comment in .tex)
 % \par
%	[\.!?][ ]?(of|in|to|for|with|on|at|from|by|about|as|into|like|through|after|over|between)[ ]([^\.!?]*[\.!?])

   \item  this, it, that, etc. replace with noun.
  \item make sure tense is consistent. youre writing about a completed piece of work so use past tense as appropriate.
  \item s/will be/..../
  \item s/will/
  \item check required format for references. if IEEE +[\#] then make sure you put references in \underline{\underline{order}} of use in dissertation.
  \item Trials - you need to beef up this section. describe all things that occurred what ??? found out from data. if anything, what could be done differently, time, issues, etc.
  \item Conclusion - restate what your dissertation contributions are.
  
  \item /s/shows/indicates/
  \item \Cref{sec:ResearchDoneToDate} - need updating, should i make this into another chapter and redo the entire section in the past tense?
  \item our - 'the' or 'the dissertation's' or 'the WHIPPED' or ''[sic]
  
  \item Show a calculation using the measured ECG and PPG trace to be uniform with your other descriptions. %great just what i wanted to do %
  \item section 3.1 "is (testing???)"
  \item Chapter 3: what else is relvant
   \begin{itemize}
   \item problems with data clarity
   \item with overwritten data
   \item lost, poorly marked,(cooelated data???)
   \item anything else
   \item talk about data you did have and how was it processed
   \item what could or should be done differently
   
   
  
   \end{itemize}
  \end{itemize}

  
\end{abstract}
